---
title: "Lab 8: Data Augmentation"
author: "Olivier Binette"
date: "Friday October 9, 2020"
fontsize: 11pt
output: 
  beamer_presentation:
    include:
      in_header: preamble.tex
---

```{r, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=4, fig.height=3, fig.align="center")
set.seed(1)
```

# Agenda

- Problem statement
- Go through the lab's tasks
- Office hours

#
\section{Problem statement}

# Problem statement

Data points $Y_1, Y_2, \dots, Y_n$ coming from a **mixture model**:
$$
  Y_i \sim^{i.i.d.} \sum_{j=1}^n w_j N(\mu_j, \varepsilon^2).
$$
\pause

**What does this mean?**\pause

For every data point:\pause

- let $Z_i$ be a random variable such that $\mathbb{P}(Z_i = j) = w_j$ for $j=1,2,3$,\pause
- let $Y_i \mid Z_i \sim N(\mu_{Z_i}, \varepsilon^2)$.

# Problem statement

Let's see what this mixture model could look like in an example.\pause

Let $\mu_1 = -5$, $\mu_2 = 0$ and $\mu_3 = 5$, and let $\varepsilon = 1$. Let $w_j = 1/3$ for $j = 1,2,3$.\pause

Now let's generate data from the mixture model:

\small
```{r, eval = FALSE}
n = 100
mu = c(-5, 0, 5)
  
Z = sample(1:3, size=n, replace=TRUE)
Y = rnorm(n, mean=mu[Z], sd=1)

hist(Y, breaks=20)
```
\normalfont



# Problem statement

```{r, eval = TRUE, echo=FALSE}
n = 100
mu = c(-5, 0, 5)
  
Z = sample(1:3, size=n, replace=TRUE)
Y = rnorm(n, mean=mu[Z], sd=1)

hist(Y, breaks=20)
```
